```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Dive into the t-test

After talking about p-values and hypothesis tests, you're probably wondering, "How do we actually test these hypotheses?" Enter the t-test, a powerful tool that helps us compare means and decide whether observed differences are statistically significant.

## Basics of the t-test

The t-test helps us determine whether two groups have different means. This test assumes that the data follows a normally distributed pattern when the sample size is small and that variances are equal, unless stated otherwise.

There are mainly two types of t-tests:
1. **Independent samples t-test:** Used when comparing the means of two separate groups, like testing a new teaching method by comparing test scores from two different classrooms.
2. **Paired sample t-test:** Useful when comparing measurements taken from the same group at different times, such as before and after a specific treatment in a medical study.

## Step-by-Step Example Using Simulated Data

Let’s consider a scenario where we're testing a new fertilizer on plant growth. We have a control group (old fertilizer) and a treatment group (new fertilizer). We want to know if the new fertilizer leads to better plant growth.

### Setting Up the Problem

```{r}
# Simulating plant heights for control and treatment groups
set.seed(42)
control <- rnorm(30, mean=20, sd=5)  # Control group, N=30, mean height = 20 cm
treatment <- rnorm(30, mean=23, sd=5)  # Treatment group, N=30, mean height = 23 cm
```

### Performing an Independent Samples t-test

```{r}
# Comparing the two groups
t_test_result <- t.test(control, treatment, alternative = "two.sided")
t_test_result
```

The output of this t-test will provide us with a p-value, which tells us if the differences in plant growth are statistically significant.

## Interpreting Results

If our p-value is less than 0.05, we can reject the null hypothesis and conclude that the new fertilizer makes a significant difference in plant growth. If it's higher, we might need more data or accept that the new fertilizer doesn't significantly outperform the old one.

Awesome! Let's tackle A/B testing next. That section will show how A/B testing is a practical application of hypothesis testing and t-tests in real-world decision-making scenarios.